First paper - Draft

### 2017-06-20 paper/paper.Rnw
Changed small thing in paper +  added all other paper docs

% How to output data frames are latex tables?
<<perc_increase_tables, cache = TRUE>>=
df <- split(
  group_by(complete_data_topmid, type_test) %>% select(wave, country, type_test, difference), complete_data_topmid$country)
.x <- df$`United States`
# Show the rates at which is increasing/decreasing
perc_increase_fun <- function(df) {

@@

>>>>

% How to output data frames are latex tables?
<<perc_increase_tables, cache = TRUE>>=
# Show the rates at which is increasing/decreasing
perc_increase_fun <- function(df) {

@@

Large inequalities in parent's education leads to large inequalities in children's education (Breen & Jonsson 2007)

The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th. (Micklewright & Schnepf, 2007)

Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (Mickleweight & Schnepf did it). They also find that France is a lower dispersion country (but in my results this is changing a lot). They also find that Germany and US are very dispersed countries (quite the opposite)

Mickleweight & Schnepf find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data. Mickleweight & Schnpef find
that average achievement is negatively related to dispersion.

Is efficiency related to opportunity here? Is efficiency what explains the inequality of opportunity?

### 2017-06-20 paper/paper.tex
Changed small thing in paper +  added all other paper docs

\section(Literature Review)

Recent research on educational inequality has found that differences in test peformance between High-SES and Low-SES kids has been growing quickly over the years. The literature on educational inequality has mainly concentrated on the United States (reardon here) but other international evidence is emerging with other similar findings for other countries. Despite this, the foremost and most important study on the subject is Reardon (2011) with American data. This is the cases, firstly, because it is the only country where cognitive testing is very widespread across surveys. And secondly, this trend has allowed to have testing records as early as 1940 (check reardon) until present day. Using this information, Reardon (2011) is the first to investigate the evolution of the cognitive gap and the results are very surprising. Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap). The widening of the achievement gap has been happening in parallel to the growth of income inequality. Although very suggestive, it is hard to link both things causally.

Many authors have taken this analysis to an international context in order to discover between-country trends. The comprehensive work of Corak and Waldfogel () shows studies, Australia, United Kingdom, United States and Canada. Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is the only study that uses panel data to study achievement gaps, let alone to do this between country}. The core finding behind the book is that the American achievement gap is much wider than in any other country. More specifically, the American achievement gap has been seen to be around ~ 1.25/1.50 standard deviations and doesn't seem to change much across the lifetime. That is, they find that the achievement gap is very stable all the way from early education to secondary school (although their studies cover up until age 14). Despite their rigurosity, the four surveys have significant differences and cannot be readily compared. This makes their results more suggestive rather than definitive. Which is why authors such as Chmielewski and Reardon (2016) and Chmielewski (2017) have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries.  Chmielewski and Reardon (2016), using the Programme for International Student Assessment (PISA), the Trends in International Mathematics and Science Study (TIMSS) and PIRLS (the Progress in International Reading Literacy Study), the assess whether there are differences in the achievement gap between countries. Although there is potential to study countries across years using these studies, given their researchn design it's impossible and only 3 countries have enough data points to be studied longitudinally. None the less, they find that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries. They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation, central exams, etc ... The correlations are indeed very suggestive but clearly we must be extremely cautios in drawing causal changes from these two variables. But it is important to stress that their design is quite different from the work of Reardon (2011). Reardon (2011) takes studies in the U.S starting from the 1940's until 2015 \footnote{Each study is independent of each other meaning that the it might've been given to 6 year olds as well as to 12 year olds. Although the author adjusts for age, the trends can't be generalized to age-specific, but rather in overall terms.} and makes them comparable across time. This gives the author a very long time series to build a reliable achievement gap (over 40 years). In their study, Chmielewski and Reardon (2016) change the aim of the study to model between-country differences from a cross-sectional perspective.

@@

Talk about reardon and portilla and how the achievement gap has been closing in the last 15 years.
Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
Talk more about how countries with high social mobility has been linked to smallest achievement gaps

\section(Research design)

\end{document}

>>>>

{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_perc_increase-1}

\section{Literature Review}

Recent research on educational inequality has found that differences in test peformance between High-SES and Low-SES kids has been growing quickly over the years. The literature on educational inequality has mainly concentrated on the United States (reardon here) but other international evidence is emerging with other similar findings for other count. The United States is usually the case study of interest, firstly, because it is the only country where cognitive testing is very widespread across surveys. And secondly, this trend has allowed to have testing records as early as 1940 (check reardon) until present day. Using this information, Reardon (2011) is the first to investigate the evolution of the cognitive gap and the results are very surprising. Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap). The widening of the achievement gap has been happening in parallel to the growth of income inequality. Although very suggestive, it is hard to link both things causally.

Reardon (2011) finds that the increase has occured predominantely from the 1970's until the 2000's. In fact, hard numbers suggest that the gap increased by about ~ 40-50\%. The author also estimates the rate of change using data from the 1940's and finds an even higher increase of about ~ 75\%. Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footenote{Even after weighting appropriately.}. Finally, Reardon (2011) concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations. Corak and Waldfogel (2015) don't find results which are very far away from this. They find that for 14 year olds in the United States, the gap is above 1 SD but lower than 1.25. Duncan and Magnuson find similarly to the previous studies and confirms a gap of ~ 1.25/1.50 standard deviations.

One important drawback of these studies is that they don't present the uncertainty of this estimate. Not necessarily to gauge their statistical significance, but to simply asses how much we can trust these numbers.

With very recent data, Reardon and Portilla (2015) have unexpectedly uncovered a new finding: the reversal of the trend. From 1998 until 2010, it seems like the income achievement gap \footnote{The income achievement gap is the difference between the 90th and 10th income percentile in the test of interest} decreased at a rate of 0.01 and 0.008 standard deviations per year. In contrast to Reardon (2011), we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something not dramatically different from the previous estimates. The American situation is a very distinctive one relative to Europe. The reasons why we're seeing a reversal in the trend could be numerous and should be studied very close to context. For example, the American educational system lacks any formal curricular differentiation \emph{à la} European style. In contrast, Reardon and Portilla (2015) suggest that the reversal is likely due to the high increase of preschool enrollment. They suggest that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.

Some authors have taken this analysis to an international context in order to discover between-country trends. The comprehensive work of Corak and Waldfogel () performs a comparative analysis of Australia, United Kingdom, United States and Canada. Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is the only study that uses panel data to study achievement gaps, let alone to do this between countries}. The core finding behind the book is that the American achievement gap is much wider than in any other comparison countries. As mentioned earlier, they find that once the achievement gap is present in early school entry, it doesn't seem to narrow very much over the life course. They find that the achievement gap is very stable all the way from early education to secondary school (although their studies cover up until age 14).

Say more above Corak and Waldfogel above (book).

Despite their high level of rigurosity, their analysis is based on four surveys that have significant differences and cannot be easily compared. Their findings are very reliable but should be taken as suggestive.  For this reason we should also pay particular attention to studies such as Chmielewski and Reardon (2016) and Chmielewski (2017) which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries. These studies tackle a completely different question from the above, but they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.

Chmielewski and Reardon (2016), using the Programme for International Student Assessment (PISA), the Trends in International Mathematics and Science Study (TIMSS) and PIRLS (the Progress in International Reading Literacy Study), the assess whether there are patterns of cross-national variation in the achievement gap. In other words, is the size of achievement gap different for many countries? Although there is potential to study countries over time using these surveys, they choose not to. They do so because their question of interest (income categories) was only asked in three waves; in the end the only had three countries repeated in all three waves.

Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries. In comparison to the literature on achievement gaps, they find that the U.S has a gap a little over 1 standard deviation in the year 2005 (for 15 year olds) and Germany has a decreasing gap from around ~ 1.25 to ~ 1 standard deviation. They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation, central exams, etc ... The correlations are indeed very suggestive but clearly we must be extremely cautios in drawing causal changes from these two variables. But it is important to stress that their design is quite different from the work of Reardon (2011). Reardon (2011) takes studies in the U.S starting from the 1940's until 2015 \footnote{Each study is independent of each other meaning that the it might've been given to 6 year olds as well as to 12 year olds. Although the author adjusts for age, the trends can't be generalized to age-specific, but rather in overall terms.} and makes them comparable across time. This gives the author a very long time series to build a reliable achievement gap (over 40 years). In their study, Chmielewski and Reardon (2016) change the aim of the study to model between-country differences from a cross-sectional perspective.

@@

Talk about reardon and portilla and how the achievement gap has been closing in the last 15 years.
Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
Talk more about how countries with high social mobility has been linked to smallest achievement gaps
Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.

\section{Research design}

Make sure PISA variation is constant over time
\end{document}
